---
title: "c3.ai COVID-19 Competition Submission"
output: html_notebook
---


Select only genes that have same mean difference direction in both data sets.
```{r}
setwd("~/Desktop/COVIDCompetition")
set.seed(2020.4)
df157103 <- read.delim('GSE157103_genes.ec.tsv',header = TRUE,sep = "\t",row.names = 1)
library(edgeR)
df157103_dgelist <- DGEList(df157103)
df157103_dgelist <- calcNormFactors(df157103_dgelist, method = 'TMM')
data1 <- cpm(df157103_dgelist)
data2 <- read.table(file = 'GSE154998_normalized_read_counts.txt', sep = '\t', header = TRUE)
colnames(data1)[1] = 'symbol'
colnames(data2)[5] = 'symbol'

GSE157103_control <- data1[,c(1,102:127)]
GSE157103_covid <- data1[,c(1:101)]
GSE157103_nonicu <- GSE157103_control[,c(1,2,5,6,10,12,22,23,24,25,26)]
GSE157103_icu <- GSE157103_control[,c(1,3,4,7,8,9,11,13,14,15,16,17,18,19,20,21,27)]
GSE157103_control_mean <- rowMeans(GSE157103_control[,2:27])
GSE157103_covid_mean <- rowMeans(GSE157103_covid[,2:101])
GSE157103_mean <- cbind(data1['symbol'],GSE157103_control_mean, GSE157103_covid_mean)

GSE154998_control <- data2[c('symbol','CO01_D1','CO09_D1','CO10_D1','CO13_D1B', 'CO19_D1', 'CO30_D1B','CO35_D1')]
GSE154998_covid <- data2[c('symbol','CO05_D1','CO25_D1X','CO40_D1','CO37_D1', 'CO11_D1', 'CO17_D1','CO22_D1')]
GSE154998_control_mean <- rowMeans(GSE154998_control[,2:8])
GSE154998_covid_mean <- rowMeans(GSE154998_covid[,2:8])
GSE154998_mean <- cbind(data2['symbol'],GSE154998_control_mean, GSE154998_covid_mean)

df_total <- merge(GSE157103_mean,GSE154998_mean,by='symbol')
df_total$GSE157103_diff <- df_total$GSE157103_control_mean - df_total$GSE157103_covid_mean
df_total$GSE154998_diff <- df_total$GSE154998_control_mean - df_total$GSE154998_covid_mean
df_sub <- df_total[df_total$GSE157103_diff * df_total$GSE154998_diff >0,]
write.csv(df_sub,"genes of the same mean difference direction.csv")
```

Select only ICU patients from GSE157103
```{r}
df <- read.delim("GSE157103_genes.tpm.tsv",header = TRUE,sep = "\t",row.names = 1)
df = as.data.frame(t(df))
info <- read.csv("patients_info.csv",header = TRUE)
info <- info[match(row.names(df),info$Albany_sampleID),]
df_icu <- df[which(info$ICU_1 ==1),]
info_icu <- info[which(info$ICU_1 ==1),]
idx = match(df_sub$symbol,names(df_icu))
subsetDf = df_icu[,idx]
subsetDf$label = replace(info_icu$COVID,which(info_icu$COVID == 1),"COVID")
subsetDf$label = replace(subsetDf$label,which(subsetDf$label == 0),"nonCOVID")
```

Create cross-validated random forest model using predictors selected from DE analysis.
```{r}
library(caret)
library(data.table)
library(randomForest)
predictionsCVRF <- matrix(0,nrow=dim(subsetDf)[1],ncol=2)
rownames(predictionsCVRF) <- rownames(subsetDf)
folds <- createFolds(subsetDf$label,10)
for(fold in folds){
  valids <- subsetDf[fold,]
  trains <- subsetDf[setdiff(1:dim(subsetDf)[1],fold),]
  
  #Perform DE Analysis on the training folds
  training_DE <- data.frame(Gene = colnames(trains), t.test=NA, p.val=NA, log2FC=NA)
  covid_samples <- grep('^C', row.names(trains))
  noncovid_samples <- grep('^NC', row.names(trains))
  n <- ncol(trains)-1
  for(i in 1:n){
    res <- t.test(trains[covid_samples,i], trains[noncovid_samples,i])
    training_DE$t.test[i] <- res$statistic
    training_DE$p.val[i] <- res$p.value
    training_DE$log2FC[i] <- log2(mean(trains[covid_samples,i])/mean(trains[noncovid_samples,i]))
  }
  training_DE$adj.p.val <- p.adjust(training_DE$p.val)
  training_DE_genes <- training_DE$Gene[which(training_DE$adj.p.val<0.05 & abs(training_DE$log2FC) > 1)]  
  df_DE_genes_names <- data.frame(training_DE_genes)
  colnames(df_DE_genes_names) <- 'symbol'
  
  trains = as.matrix.data.frame(trains)
  trains = data.frame(t(trains))
  setDT(trains, keep.rownames = TRUE)[]
  colnames(trains)[1] = 'symbol'
  geneDESubset <- merge(trains,df_DE_genes_names,by='symbol')
  trains = as.data.frame(t(geneDESubset))
  names(trains) = as.matrix(trains[1,])
  trains = trains[-1,]
  trains[] <- lapply(trains,function(x) as.numeric(as.character(x)))
  trains$label = 0
  trains[covid_samples,]$label = "COVID"
  trains[noncovid_samples,]$label = "nonCOVID"
  trains$label <- as.factor(trains$label)

  valids = as.matrix.data.frame(valids)
  valids = data.frame(t(valids))
  setDT(valids, keep.rownames = TRUE)[]
  colnames(valids)[1] = 'symbol'
  geneDESubsetValids <- merge(valids,df_DE_genes_names,by='symbol')
  valids = as.data.frame(t(geneDESubsetValids))
  names(valids) = as.matrix(valids[1,])
  valids = valids[-1,]
  valids[] <- lapply(valids,function(x) as.numeric(as.character(x)))
  valids$label = 0
  covid_samples_valids <- grep('^C', row.names(valids))
  noncovid_samples_valids <- grep('^NC', row.names(valids))
  valids[covid_samples_valids,]$label = "COVID"
  valids[noncovid_samples_valids,]$label = "nonCOVID"
  valids$label <- as.factor(valids$label)

  #Create a random forest model using the predictors from DE analysis

  tmpRF <- randomForest(label ~ . ,data=trains,importance=T,ntree=1000,nodesize=5)
  predicted <- predict(tmpRF,valids,type='prob')
  predictionsCVRF[rownames(predicted),] <- predictionsCVRF[rownames(predicted),]+predicted
}
colnames(predictionsCVRF) <- colnames(predicted)

predicts <- t(apply(predictionsCVRF,1,function(v){v/sum(v)}))
colnames(predicts) <- colnames(predicted)
predicts <- data.frame(predicts,check.names=F)
predicts$predicted <- apply(predicts,1,function(v){names(v)[max(v)==v]})
predicts$observed <- subsetDf$label
predicts
```

Accuracy of Random Forest Cross Validation
```{r}
accuracy <- sum(as.matrix(predicts$predicted == predicts$observed))
accuracy/nrow(predicts)
tp = sum(as.matrix(predicts$predicted == "COVID" & predicts$observed == "COVID"))
tn = sum(as.matrix(predicts$predicted == "nonCOVID" & predicts$observed == "nonCOVID"))
fn = sum(as.matrix(predicts$predicted == "nonCOVID" & predicts$observed == "COVID"))
fp = sum(as.matrix(predicts$predicted == "COVID" & predicts$observed == "nonCOVID"))

predicts$p
```

Create cross-validated xgboost model using predictors selected from DE analysis.
```{r}
library(xgboost)
predictionsXGBCV <- matrix(0,nrow=dim(subsetDf)[1],ncol=2)
rownames(predictionsXGBCV) <- rownames(subsetDf)
folds <- createFolds(subsetDf$label,10)
for(fold in folds){
  valids <- subsetDf[fold,]
  trains <- subsetDf[setdiff(1:dim(subsetDf)[1],fold),]
  
  #Perform DE Analysis on the training folds
  training_DE <- data.frame(Gene = colnames(trains), t.test=NA, p.val=NA, log2FC=NA)
  covid_samples <- grep('^C', row.names(trains))
  noncovid_samples <- grep('^NC', row.names(trains))
  n <- ncol(trains)-1
  for(i in 1:n){
    res <- t.test(trains[covid_samples,i], trains[noncovid_samples,i])
    training_DE$t.test[i] <- res$statistic
    training_DE$p.val[i] <- res$p.value
    training_DE$log2FC[i] <- log2(mean(trains[covid_samples,i])/mean(trains[noncovid_samples,i]))
  }
  training_DE$adj.p.val <- p.adjust(training_DE$p.val)
  training_DE_genes <- training_DE$Gene[which(training_DE$adj.p.val<0.05 & abs(training_DE$log2FC) > 1)]  
  df_DE_genes_names <- data.frame(training_DE_genes)
  colnames(df_DE_genes_names) <- 'symbol'
  
  trains = as.matrix.data.frame(trains)
  trains = data.frame(t(trains))
  setDT(trains, keep.rownames = TRUE)[]
  colnames(trains)[1] = 'symbol'
  geneDESubset <- merge(trains,df_DE_genes_names,by='symbol')
  trains = as.data.frame(t(geneDESubset))
  names(trains) = as.matrix(trains[1,])
  trains = trains[-1,]
  trains[] <- lapply(trains,function(x) as.numeric(as.character(x)))
  trains$label = 0
  trains[covid_samples,]$label = "COVID"
  trains[noncovid_samples,]$label = "nonCOVID"
  trains$label <- as.factor(trains$label)

  valids = as.matrix.data.frame(valids)
  valids = data.frame(t(valids))
  setDT(valids, keep.rownames = TRUE)[]
  colnames(valids)[1] = 'symbol'
  geneDESubsetValids <- merge(valids,df_DE_genes_names,by='symbol')
  valids = as.data.frame(t(geneDESubsetValids))
  names(valids) = as.matrix(valids[1,])
  valids = valids[-1,]
  valids[] <- lapply(valids,function(x) as.numeric(as.character(x)))
  valids$label = 0
  covid_samples_valids <- grep('^C', row.names(valids))
  noncovid_samples_valids <- grep('^NC', row.names(valids))
  valids[covid_samples_valids,]$label = "COVID"
  valids[noncovid_samples_valids,]$label = "nonCOVID"
  valids$label <- as.factor(valids$label)

  #Create a xgb model using the predictors from DE analysis
  dataTrainMatrix = data.matrix(trains[, colnames(trains) != "label" ])
  labelTrainMatrix = data.matrix(as.numeric(trains$label == "COVID"))
  dataValidationMatrix = data.matrix(valids[, colnames(valids) != "label" ])
	tmpXGB <- xgboost(data=dataTrainMatrix,label=labelTrainMatrix, objective="binary:logistic",nrounds=5)
  predicted <- predict(tmpXGB,dataValidationMatrix,type='prob')
  predictionsXGBCV[rownames(valids),] <- predictionsXGBCV[rownames(valids),]+predicted
}
predictionsXGBCV = as.data.frame(predictionsXGBCV)
colnames(predictionsXGBCV) <- c("VAL1","VAL2")
predictionsXGBCV$Observed = subsetDf$label
predictionsXGBCV$Predicted = lapply(predictionsXGBCV$VAL1, function(x) ifelse(x>0.5, "COVID","nonCOVID"))

predictionsXGBCV
```


Accuracy of xgboost Cross Validation
```{r}
accuracy <- sum(as.matrix(predictionsXGBCV$Predicted == predictionsXGBCV$Observed))
accuracy/nrow(predictionsXGBCV)
tp = sum(as.matrix(predictionsXGBCV$Predicted == "COVID" & predictionsXGBCV$Observed == "COVID"))
tn = sum(as.matrix(predictionsXGBCV$Predicted == "nonCOVID" & predictionsXGBCV$Observed == "nonCOVID"))
fn = sum(as.matrix(predictionsXGBCV$Predicted == "nonCOVID" & predictionsXGBCV$Observed == "COVID"))
fp = sum(as.matrix(predictionsXGBCV$Predicted == "COVID" & predictionsXGBCV$Observed == "nonCOVID"))
```

We have established that both models generalize well from training to testing data. Now, we will test how well a model trained from a calibrated data set performs on testing data both from the same calibrated data set and the data set it has been calibrated with.


Perform DE Analysis on Training Set
80% of GSE157103 will be used as training, 20% as testing
```{r}
library(caret)
split <- createDataPartition(subsetDf$label,list=FALSE, p=0.8)
training <- subsetDf[split,]
test <- subsetDf[-split,]

training_DE <- data.frame(Gene = colnames(training), t.test=NA, p.val=NA, log2FC=NA)
covid_samples <- grep('^C', row.names(training))
noncovid_samples <- grep('^NC', row.names(training))
n <- ncol(training)-1
for(i in 1:n){
  res <- t.test(training[covid_samples,i], training[noncovid_samples,i])
  training_DE$t.test[i] <- res$statistic
  training_DE$p.val[i] <- res$p.value
  training_DE$log2FC[i] <- log2(mean(training[covid_samples,i])/mean(training[noncovid_samples,i]))
}
training_DE$adj.p.val <- p.adjust(training_DE$p.val)
training_DE_genes <- training_DE$Gene[which(training_DE$adj.p.val<0.05 & abs(training_DE$log2FC) > 1)]

```

Perform log2 transformation 
```{r}
df_DE_genes_names <- data.frame(training_DE_genes)
colnames(df_DE_genes_names) <- 'symbol'
geneDESubset <- merge(df_sub,df_DE_genes_names,by='symbol')

GSE157103_control_sub <- merge(GSE157103_icu,geneDESubset['symbol'],by='symbol')
GSE154998_control_sub <- merge(GSE154998_control,geneDESubset['symbol'],by='symbol')
trans157103DE <- geneDESubset['symbol']
for (i in seq(2,17)){
  v1 <- log2(GSE157103_control_sub[,i]+1)
  trans157103DE <- cbind(trans157103DE,v1)
}
colnames(trans157103DE) <- colnames(GSE157103_icu)

trans154998DE <- geneDESubset['symbol']
for (i in seq(2,8)){
  v2 <- log2(GSE154998_control_sub[,i]+1)
  trans154998DE <- cbind(trans154998DE,v2)
}
colnames(trans154998DE) <- colnames(GSE154998_control)
```

Calibration using Geometric Mean Regression (GMR)
```{r}
### GMR equation: GSE157103 = slope * GSE154998 + intercept
log2_GSE157103_mean <- rowMeans(trans157103DE[,2:17])
log2_GSE154998_mean <- rowMeans(trans154998DE[,2:8])
slope <- sd(log2_GSE157103_mean) / sd(log2_GSE154998_mean)
intercept <- mean(log2_GSE157103_mean) - slope * mean(log2_GSE154998_mean)

ggplot(data.frame(log2_GSE157103_mean,log2_GSE154998_mean), aes(x = log2_GSE154998_mean, y = log2_GSE157103_mean))+
  geom_point()+
  geom_abline(intercept = intercept, slope = slope)+
  geom_abline(intercept = 0, slope = 1, color="blue",linetype='dashed')+
  labs(title="Scatter plot of gene mean expression")+
  geom_text(x = 6.5, y = 7.5, label = "y = 1.004x + 0.137") + 
  geom_text(x = 7.5, y = 6, label = "y = x", color='blue')+
  theme_classic()
ggsave("GMR_fit.jpeg")

GSE154998 <- merge(data2[,-c(1:4)],geneDESubset['symbol'],by='symbol')
GSE154998_calibrated <- geneDESubset['symbol']
for (i in colnames(GSE154998[-1])){
  temp <- 2 ^ (slope * log2(GSE154998[i]+1) + intercept) - 1
  GSE154998_calibrated <- cbind(GSE154998_calibrated,temp)
}
GSE157103 <- merge(data1,geneDESubset['symbol'],by='symbol')
GSE157103_training = GSE157103[,split]
GSE157103_test = GSE157103[,-split]
write.csv(GSE157103_training,'GSE157103_training.csv')
write.csv(GSE157103_test,'GSE157103_test.csv')
write.csv(GSE154998_calibrated,'GSE154998_calibrated.csv')
```

Training a random forest model with calibrated GSE157103 training set
```{r}
library(caret)
library(data.table)
library(randomForest)

GSE157103_calibrated = as.data.frame(t(GSE157103_calibrated))
names(GSE157103_calibrated) = as.matrix(GSE157103_calibrated[1,])
GSE157103_calibrated = GSE157103_calibrated[-1,]
GSE157103_calibrated[] <- lapply(GSE157103_calibrated,function(x) as.numeric(as.character(x)))
info <- read.csv("patients_info.csv",header = TRUE)
info <- info[match(row.names(GSE157103_calibrated),info$Albany_sampleID),]
df_icu <- GSE157103_calibrated[which(info$ICU_1 ==1),]
info_icu <- info[which(info$ICU_1 ==1),]
df_icu$label = replace(info_icu$COVID,which(info_icu$COVID == 1),"COVID")
df_icu$label = replace(df_icu$label,which(df_icu$label == 0),"nonCOVID")
df_icu$label = as.factor(df_icu$label)
training = df_icu[split,]
test = df_icu[-split,]

rfModel <- randomForest(label ~ . ,data=training,importance=T,ntree=1000,nodesize=5)
predictedGSE157103Test <- predict(rfModel,test,type='prob')
predictedGSE157103Test = as.data.frame(predictedGSE157103Test)
predictedGSE157103Test$predicted <- lapply(predictedGSE157103Test$COVID, function(x) ifelse(x>0.5, "COVID","nonCOVID"))
predictedGSE157103Test$observed <- test$label
predictedGSE157103Test
importance = data.frame(rfModel$importance[,4])
colnames(importance) = c("MeanDecreaseAccuracy")
```

Accuracy of Calibrated Random Forest Model on Test Data from same Data Set
```{r}
accuracy <- sum(as.matrix(predictedGSE157103Test$predicted == predictedGSE157103Test$observed))
accuracy/nrow(predictedGSE157103Test)
tp = sum(as.matrix(predictedGSE157103Test$predicted == "COVID" & predictedGSE157103Test$observed == "COVID"))
tn = sum(as.matrix(predictedGSE157103Test$predicted == "nonCOVID" & predictedGSE157103Test$observed == "nonCOVID"))
fn = sum(as.matrix(predictedGSE157103Test$predicted == "nonCOVID" & predictedGSE157103Test$observed == "COVID"))
fp = sum(as.matrix(predictedGSE157103Test$predicted == "COVID" & predictedGSE157103Test$observed == "nonCOVID"))
```

Testing Accuracy on Different Data Set: GSE154998
```{r}
controlDF2 = as.data.frame(t(GSE154998_control))
names(controlDF2) = as.matrix(controlDF2[1,])
controlDF2 = controlDF2[-1,]
controlDF2[] <- lapply(controlDF2,function(x) as.numeric(as.character(x)))
controlDF2$label = "nonCOVID"

covidDF2 = as.data.frame(t(GSE154998_covid))
names(covidDF2) = as.matrix(covidDF2[1,])
covidDF2 = covidDF2[-1,]
covidDF2[] <- lapply(covidDF2,function(x) as.numeric(as.character(x)))
covidDF2$label = "COVID"

testDF2 = rbind(covidDF2,controlDF2)


predictedGSE154998 <- predict(rfModel,testDF2,type='prob')
predictedGSE154998 = as.data.frame(predictedGSE154998)
predictedGSE154998$predicted <- lapply(predictedGSE154998$COVID, function(x) ifelse(x>0.5, "COVID","nonCOVID"))
predictedGSE154998$observed <- testDF2$label
predictedGSE154998
```

Accuracy of Calibrated Random Forest Model on Test Data from Separate Calibrated Data Set
```{r}
accuracy <- sum(as.matrix(predictedGSE154998$predicted == predictedGSE154998$observed))
accuracy/nrow(predictedGSE154998)
tp = sum(as.matrix(predictedGSE154998$predicted == "COVID" & predictedGSE154998$observed == "COVID"))
tn = sum(as.matrix(predictedGSE154998$predicted == "nonCOVID" & predictedGSE154998$observed == "nonCOVID"))
fn = sum(as.matrix(predictedGSE154998$predicted == "nonCOVID" & predictedGSE154998$observed == "COVID"))
fp = sum(as.matrix(predictedGSE154998$predicted == "COVID" & predictedGSE154998$observed == "nonCOVID"))
```

Training an xgboost model with calibrated GSE157103 training set
```{r}

library(xgboost)

training = df_icu[split,]
test = df_icu[-split,]
dataTrainMatrix = data.matrix(training[, colnames(training) != "label" ])
labelTrainMatrix = data.matrix(as.numeric(training$label == "COVID"))
dataTestGSE157103Matrix = data.matrix(test[, colnames(test) != "label" ])
xgbModel <- xgboost(data=dataTrainMatrix,label=labelTrainMatrix, objective="binary:logistic",nrounds=5)

predictedGSE157103TestXGB <- predict(xgbModel,dataTestGSE157103Matrix,type='prob')
predictedGSE157103TestXGB = as.data.frame(predictedGSE157103TestXGB)
predictedGSE157103TestXGB$predicted <- lapply(predictedGSE157103TestXGB[,1], function(x) ifelse(x>0.5, "COVID","nonCOVID"))
predictedGSE157103TestXGB$observed <- test$label
predictedGSE157103TestXGB
```

Accuracy of Calibrated xgboost Model on Test Data from same Data Set
```{r}
accuracy <- sum(as.matrix(predictedGSE157103TestXGB$predicted == predictedGSE157103TestXGB$observed))
accuracy/nrow(predictedGSE157103TestXGB)
tp = sum(as.matrix(predictedGSE157103TestXGB$predicted == "COVID" & predictedGSE157103TestXGB$observed == "COVID"))
tn = sum(as.matrix(predictedGSE157103TestXGB$predicted == "nonCOVID" & predictedGSE157103TestXGB$observed == "nonCOVID"))
fn = sum(as.matrix(predictedGSE157103TestXGB$predicted == "nonCOVID" & predictedGSE157103TestXGB$observed == "COVID"))
fp = sum(as.matrix(predictedGSE157103TestXGB$predicted == "COVID" & predictedGSE157103TestXGB$observed == "nonCOVID"))
```

Testing Accuracy of xgboost model on Different Data Set: GSE154998
```{r}

dataTestGSE154998 = data.matrix(testDF2[,names(GSE157103_calibrated)])
predictedGSE154998XGB <- predict(xgbModel,dataTestGSE154998,type='prob')
predictedGSE154998XGB = as.data.frame(predictedGSE154998XGB)
predictedGSE154998XGB$predicted <- lapply(predictedGSE154998XGB[,1], function(x) ifelse(x>0.5, "COVID","nonCOVID"))
predictedGSE154998XGB$observed <- testDF2$label
predictedGSE154998XGB
```

Accuracy of Calibrated xgboost Model on Test Data from Calibrated Second Data Set
```{r}
accuracy <- sum(as.matrix(predictedGSE154998XGB$predicted == predictedGSE154998XGB$observed))
accuracy/nrow(predictedGSE154998XGB)
tp = sum(as.matrix(predictedGSE154998XGB$predicted == "COVID" & predictedGSE154998XGB$observed == "COVID"))
tn = sum(as.matrix(predictedGSE154998XGB$predicted == "nonCOVID" & predictedGSE154998XGB$observed == "nonCOVID"))
fn = sum(as.matrix(predictedGSE154998XGB$predicted == "nonCOVID" & predictedGSE154998XGB$observed == "COVID"))
fp = sum(as.matrix(predictedGSE154998XGB$predicted == "COVID" & predictedGSE154998XGB$observed == "nonCOVID"))
```
